---
title: "Bellabeat Case Study"
output: html_document
date: "2024-11-27"
author: Samantha Schloeman
---
## The Business Task 
#### In this case study, we're tasked with analyzing fitbit user data from the FitBit Fitness Tracker dataset found on Kaggle: <https://www.kaggle.com/datasets/arashnic/fitbit> (CC0: Public Domain, dataset made available through Mobius).

#### Our goal is to understand user behavior in order to supply the Bellabeat marketing team with a set of recommendations to grow one of the brand's product offerings. In this case, I chose to focus on the brand's wellness application, the Bellabeat App. 

#### The questions underlying my analysis include: 
-What trends exist in smart device usage?   
-How could these trends apply to Bellabeat customers?   
-How could these trends influence the Bellabeat marketing strategy?   

## Download and Inspect the Data
#### Setting up the environment
```{r, message=FALSE, results='hide'}
#Load libraries
library(readr)  
library(janitor)  
library(here)
library(skimr)
library(dplyr)  
library(lubridate)  
library(ggplot2) 
library(tidyr)
library(Hmisc)

#read files
dailyactivity_merged_marapr <- read_csv("dailyActivity_merged.csv")
dailyactivity_merged_aprmay <- read_csv("dailyActivity_merged_AprMay.csv")
weightmarapr <- read_csv("weightLogInfo_merged_MarApr.csv")
weightaprmay <- read_csv("weightLogInfo_merged_AprMay.csv")
```


```{r}
#Review
head(dailyactivity_merged_marapr)
head(dailyactivity_merged_aprmay)
head(weightmarapr)
head(weightaprmay)

skim_without_charts(dailyactivity_merged_marapr)
skim_without_charts(dailyactivity_merged_aprmay)
skim_without_charts(weightmarapr)
skim_without_charts(weightaprmay)
```
#### The unique user Id column will be useful. 


#### Synchronize the date variables. Remove duplicates, unwanted columns, and compare column names to merge the like dataframes. 

```{r}
#Remove duplicates
dailyactivity_merged_marapr <- distinct(dailyactivity_merged_marapr)
dailyactivity_merged_aprmay <- distinct(dailyactivity_merged_aprmay)
weightmarapr <- distinct(weightmarapr)
weightaprmay <- distinct(weightaprmay)

#Remove unwanted columns
weightmarapr <- weightmarapr %>%
  select(-Fat, -BMI, -WeightKg)
weightaprmay <- weightaprmay %>%
  select(-Fat, -BMI, -WeightKg)

#Compare col names for the rbind
compare_df_cols_same(dailyactivity_merged_marapr, dailyactivity_merged_aprmay)
compare_df_cols_same(weightmarapr, weightaprmay)

#merge like dataframes 
daily_activity <- rbind(dailyactivity_merged_marapr, dailyactivity_merged_aprmay)
daily_weight <- rbind(weightmarapr, weightaprmay)

#Convert character to date in the activity df
daily_activity <- daily_activity %>%
  mutate(ActivityDate = convert_to_date(ActivityDate, 
                                        character_fun = lubridate::mdy))

#Isolate date from time in the weight df
daily_weight <- daily_weight %>%
  separate(Date, c("Day", NA), " ", remove = FALSE, extra = "merge") %>%
  mutate(Day = convert_to_date(Day,
                                character_fun = lubridate::mdy))
```

#### How many unique participants are in each df? 
```{r}
n_distinct(daily_activity$Id) 
n_distinct(daily_weight$Id) 
```

#### We'll focus on sedentary minutes, very active minutes, calories, and weight in this analysis. More time spent sedentary is correlated with adverse health outcomes according to [the NIH.](https://pmc.ncbi.nlm.nih.gov/articles/PMC6527341/#:~:text=Results:,the%202018%20PAGAC%20Scientific%20Report.) 

#### Summary statistics
```{r}
daily_activity %>%
  select(TotalSteps, TotalDistance, SedentaryMinutes) %>%
  summary()

daily_weight %>%
  select(WeightPounds, IsManualReport) %>%
  summary()
```
#### 75% of our participants log less than 11k steps per day and spend more than 729 minutes (>12 hours) sedentary. 75% are less than 188 lbs and 2/3 of the weight records are manually entered. 

#### How do sedentary minutes change throughout the week? 
```{r}
daily_activity$Weekday <- daily_activity$ActivityDate %>% weekdays()

daily_activity %>% 
  group_by(Weekday) %>% 
  summarise(mean_sedentary_mins = mean(SedentaryMinutes)) %>%
  arrange(mean_sedentary_mins) %>%
  ggplot(aes(x=Weekday, y=mean_sedentary_mins, label=round(mean_sedentary_mins), fill = "ef8d76")) + 
  geom_col() +
  geom_text(position = "identity") +
  geom_hline(aes(yintercept = mean(mean_sedentary_mins))) +
  labs(y="Mean Sedentary Mins") + 
  theme_bw() +
  theme(legend.position = "none")
```

#### How does weight change through time? 
```{r}
daily_weight %>%
ggplot(aes(x=Day, y=WeightPounds, color=IsManualReport)) +
  geom_line() + 
  geom_point() +  
  facet_wrap(~Id) +
  labs(y="Weight (lbs)", x="Date", color="Is Manual Report?") +
  theme_bw() +
  theme(strip.background = element_rect("white"), legend.position = "bottom") +
  scale_color_manual(values=c("#f5c885", "#e94472")) 
```

#### Only 2 out of 13 users are tracking their weight consistently. 

#### What's the relationship between total steps and calories? 
```{r}
ggplot(data = daily_activity, aes(x=TotalSteps, y=Calories, color="#e94472")) +
  geom_point() +
  geom_smooth() +
  labs(x="Total Steps") +
  theme_bw() +
  theme(legend.position = "none")

cor(daily_activity$TotalSteps, daily_activity$Calories)
```
#### Surprisingly, there's only a weak correlation between total steps and calories burned. 

#### How does logged distance differ from total distance? 
```{r, warning=FALSE}
daily_activity$LoggedActivitiesDistance %>%
  summary()

daily_activity$TotalDistance %>%
  summary()

daily_activity$LoggedPercentage <- ((daily_activity$LoggedActivitiesDistance/daily_activity$TotalDistance)*100)
summary(daily_activity$LoggedPercentage)

daily_activity %>%
  na.omit() %>%
  ggplot(aes(x=LoggedPercentage, fill = "#f5c885")) + 
  geom_area(stat="bin") +
  labs(x="Logged Total Activity Distance (%)", y="Count of Days") +
  theme_bw() +
  theme(legend.position = "none")
```

#### Do sedentary minutes change through time spent with the device? 
```{r}
daily_activity %>%
  select(Id, ActivityDate, SedentaryMinutes) %>%
  ggplot(aes(x=ActivityDate, y=SedentaryMinutes, color=Id)) +
  geom_point(position="jitter") +
  theme_bw() +
  labs(x="Date", y="Sedentary Minutes")
```

#### Let's look at a small sample of users. 
```{r}
sed_avg_wks1and2 <- daily_activity %>%
  filter(ActivityDate < "2016-03-26") %>%
  group_by(Id) %>%
  summarise(mean_sed_mins_first = mean(SedentaryMinutes))
sed_avg_wks1and2 

sed_avg_lastwk <- daily_activity %>%
  filter(ActivityDate < "2016-05-12") %>%
  group_by(Id) %>%
  summarise(mean_sed_mins_last = mean(SedentaryMinutes))
sed_avg_lastwk

sed_avg_wks1and2$Id %in% sed_avg_lastwk$Id #Our 4 participants are in the last week 
sed_time_comp <- inner_join(sed_avg_wks1and2, sed_avg_lastwk, by="Id")
sed_time_comp %>% 
  mutate(change = mean_sed_mins_last - mean_sed_mins_first)
```
#### 3 of 4 users decreased their sedentary time between the first 2 weeks and the last week. This is an interesting question to explore with long-term data and more users. 

#### How do sedentary minutes change with very active minutes?
```{r, warning=FALSE}
ggplot(data=daily_activity, aes(x=VeryActiveMinutes, y=SedentaryMinutes, color=Id)) +
  geom_point() +
  theme_bw() +
  labs(x="Very Active Minutes", y="Sedentary Minutes")

#Take a sample of participants
user_sample <- sample(daily_activity$Id, 5)

daily_activity %>%
  filter(Id %in% user_sample) %>% 
  filter(SedentaryMinutes > 0, VeryActiveMinutes > 0) %>%
  ggplot(aes(x=VeryActiveMinutes, y=SedentaryMinutes, group=Id, color=Id)) +
  geom_point() + 
  geom_smooth() +
  theme_bw() +
  labs(x="Very Active Minutes", y="Sedentary Minutes")

cor(daily_activity$VeryActiveMinutes, daily_activity$SedentaryMinutes) 
```
#### There is no evidence that a correlation exists between very active and sedentary minutes. 

#### Merge the activity and weight dataframes. 
```{r}
names(daily_activity)[2] <- "Day"
activity_weight_merge <- merge(daily_activity, daily_weight, by=c("Id", "Day"))
n_distinct(activity_weight_merge$Id) #13
```
#### There are 13 unique users after the merge. 

#### Is there any correlation between calories, sedentary time, very active time, and weight? 
```{r}
correlation_test <- activity_weight_merge %>%
  select(Calories, SedentaryMinutes, VeryActiveMinutes, WeightPounds) %>%
  as.matrix()
rcorr(correlation_test) 
#Strong positive correlation between Calories and VeryActiveMinutes
#Slight positive corrlation between SedentaryMinutes and Calories, WeightPounds and Calories
#No correlation between VeryActiveMinutes and SedentaryMinutes, WeightPounds and VeryActiveMinutes

#Calories and VeryActiveMinutes 
activity_weight_merge %>%
  select(Id, Calories, VeryActiveMinutes, Day) %>%
  arrange(Day) %>%
  ggplot(aes(x=Calories, y=VeryActiveMinutes, color=Id)) +
  geom_point() +
  theme_bw() +
  labs(y="Very Active Minutes")
```

## In Summary 
### We downloaded, cleaned, and performed exploratory data analysis on FitBit User data to better understand user behavior. 
#### The data tells us that:
1. Most users spend their days sedentary. Average sedentary time is above the recommended maximum of 8 hours per day. 
2. Users do not track their weight consistently. 
3. 3 out of 4 users suitable for comparison decreased their sedentary minutes over time. 
4. Users do not log their activity often. 
5. Surprisingly, calories and total steps are not strongly correlated. Neither are sedentary minutes and very active minutes. There is a strong positive correlation between very active minutes and calories. 

#### These findings suggest that:
1. Users may struggle to schedule activity breaks throughout their day.
2. Weight loss may not be the primary goal for most users, as they are not consistently tracking their weight. 
3. Users may be motivated to spend less time sedentary when they wear their devices. 

### Suggestions for the Bellabeat App marketing strategy
1. Focus development resources on creating more short form moderate to very high intensity workouts available through the app to encourage users to break up long periods of sitting. 
2. Survey Bellabeat users about their wellness goals. Create a feature that allows users to disable weight and calorie tracking if they don't find these features useful to their wellness journey. 
3. Perform a long-term study to explore the relationship between device use and decreased sedentary time. Highlight this benefit in marketing materials. 
4. Align marketing strategies, including brand ambassadors and partners, with user shift towards holistic wellness. 

